---
title: "CompBio Hackathon 2026: exploring fusion genes in cancer"
subtitle: "TCGAbiolinks & cbioportalR to access TCGA data"
author: "Beto Bermudez & Robin Shaw"
date-modified: last-modified
date-format: "D MMM, YYYY"
abstract: "Exploring fusion genes in TCGA. Here we pull public informatin from fusion genes and clinical data"
title-block-banner: true

execute:
  echo: true
  error: false
  warning: false
  cache: false

format:
  html: 
    grid:
      sidebar-width: 200px
      body-width: 2000px
      margin-width: 400px
      gutter-width: 1.5rem
    page-layout: full
    toc: true
    toc-depth: 2
    toc-location: "left"
    toc-title: "Contents"
    cap-location: "bottom"
    fig-cap-location: bottom
    code-fold: true
    code-overflow: "wrap"
    code-copy: true
    self-contained: true
    embed-resources: true
    number_sections: true

editor: visual
---

# Introduction

--Details of the project and any previous analysis--

## Packages and R

The analysis will be computed using the R Language [@rcoreteam2020] and using packages from CRAN and Bioconductor [@huber2015].

```{r }
#| label: libraries
suppressMessages({
  if(!require(tidyverse)) install.packages("tidyverse")
  
  if(!require(TCGAbiolinks)) install.packages("TCGAbiolinks")
 
  if(!require(dplyr)) install.packages("dplyr")
  if(!require(tidyr)) install.packages("tidyr")
  
  if(!require(gt)) install.packages("gt")
  if(!require(Cairo)) install.packages("Cairo")
})

```

```{r}
#| label: main_parameters
species <- "Human"
annot_version <- 114
```

# Get annotation data

The Ensembl Biomart database [@cunningham2022] is used to get annotation data.

::: {.callout-note collapse="true"}
For older reference versions of ensemble use `listEnsemblArchives()` to find the relevant mart.

Get information on:

-   ensembl_gene_id for matching
-   mgi_symbol/external_gene_name for identifiable gene name
-   description - Gene description
-   chromosome_name - Chromosome/scaffold name
-   start_position - Gene start (bp)
-   end_position - Gene end (bp)
-   entrezgene - NCBI gene
:::

```{r eval=FALSE}
#| label: tbl-marts
#| tbl-cap: "Available marts"
# Get gene names from biomaRt
# a random bug is arised from listMarts(), https://support.bioconductor.org/p/9154901/
listMarts(host="http://www.ensembl.org") |> gt() |> opt_stylize(style = 6, color = 'gray') # http for HPC
```

```{r eval=FALSE}
#| label: tbl-legacy
#| tbl-cap: "Last 5 Archived References"
head(listEnsemblArchives()) |> gt() |> opt_stylize(style = 6, color = 'gray') 
```

```{r get_annotation}
#| label: tbl-get_annotation
#| tbl-cap: "Table of Annotation (head of table)"
#reference_dir <- "~/data/Y90/References_Index/references" # local
reference_dir <- "~/shared_data/references_index" # for HPC
biomart_version <- annot_version

if (species == "Mouse" & biomart_version >= 103) {
  biomart_version <- c(39, biomart_version)
} else {
  biomart_version <- c(38, biomart_version)
}

# Read gene annotation
biomart_annotation_file <- ifelse(species == "Mouse",
  file.path(reference_dir,"Mouse",paste0("version_",biomart_version[1],".",biomart_version[2]),paste0("biomaRt_Mouse_",biomart_version[1],"_", biomart_version[2],".rds")),
  file.path(reference_dir,"Human",paste0("Homo_sapiens_",biomart_version[1],".",biomart_version[2]),paste0("biomaRt_Human_",biomart_version[1],"_", biomart_version[2],".rds")))
if (!file.exists(biomart_annotation_file)) {
  # Get gene names from biomaRt
  biomaRt::listMarts(host="http://www.ensembl.org") # http for HPC
  
  #Get list of archived data
  head(biomaRt::listEnsemblArchives())

  attributes_ <- c('ensembl_gene_id','description','external_gene_name', "entrezgene_id", "gene_biotype",
                  "chromosome_name", 'start_position', 'end_position')
  biomart_host <- biomaRt::listEnsemblArchives()
  ensembl <- biomaRt::useMart(biomart = 'ENSEMBL_MART_ENSEMBL', 
                     dataset = ifelse(species == "Mouse", 'mmusculus_gene_ensembl', 'hsapiens_gene_ensembl'),
                     host = biomart_host$url[match(biomart_version[2], biomart_host$version)]
                    )
  annot_raw <- biomaRt::getBM(
      attributes = attributes_,
      mart = ensembl)
  saveRDS(annot_raw, file = biomart_annotation_file)
} else {
  annot_raw <- readRDS(biomart_annotation_file)
  annot_raw$description = sub(" \\[Source.*\\]$", "", annot_raw$description)
}

# no duplication?
# stopifnot(any(duplicated(annot_raw$ensembl_gene_id)) == FALSE)
# table(duplicated(annot_raw$ensembl_gene_id))
# There are duplication due to multiple entrezgene_id

# Save the complete annotation data
#data.table::fwrite(annot_raw, paste0(exper_id, "_annotation.tsv"), sep = "\t")

# Only keep unique values
annot <- annot_raw[!duplicated(annot_raw$ensembl_gene_id),]
tail(annot) |> gt() |> opt_stylize(style = 6, color = 'gray') 
```

# TCGA-LUAD Sequencing Datas

Genomic bam for 00d461ae-a1d8-42f2-abd8-5e159363d857 is 11 Gb

transcriptome bam for 00d461ae-a1d8-42f2-abd8-5e159363d857 is \~6.5 Gb

chimeric reads fam for 00d461ae-a1d8-42f2-abd8-5e159363d857 is \~60 Mb.

```{r}
df <- read.table("~/long-term-scratch/bbermude/data/OMICS/Y90_2409_JRBB_Test/Information/gdc_manifest.txt", header = T)

head(df %>% arrange(filename), n = 10) |> gt() |> opt_stylize(style = 6, color = 'gray') 
```

```{r}
df$type <- sub("^[^.]*\\.", "", df$filename)


df$type <- ifelse(df$type == "bam", yes = sub("^[^_]*_", "", df$filename), no = df$type)

df$type <- sub(".bam$", "", df$type)

df$type <- sub("gdc_realn$", "", df$type)

df$type <- sub("\\.$", "", df$type)
df$type <- sub("_$", "", df$type)
```

I think that:

-   wxs: whole exome sequencing
-   WholeGenome.RP-1657: ???
-   wgs: whole genome sequencing
-   rna_seq.chimeric: ???
-   rna_seq.genomic: ???
-   rna_seq.transcriptome: bulk RNA-seq
-   mirna: small RNAs sequencing.
-   atacseq: Assay for Transposase-Accessible Chromatin using sequencing.

Interestingly the number of rna_seq.chimeric, rna_seq.genomic and rna_seq.transcriptome matches perfectly.

```{r}
sort(table(df$type), decreasing = T)
```

```{r}
barplot(sort(table(df$type), decreasing = T), las=2, main="most abundant data types for LUAD")
```

The biggest files are wgs and WholeGenome.RP-1657

```{r}
l <- split(df$size, df$type)
boxplot(l, main="file size by type", las=2)
```

We won't be working with WGS data during the hackathon, drop these entries from the table.

```{r }
df <- df %>% filter(type != "wgs") %>% filter(type != "WholeGenome.RP-1657")
```

```{r}
l <- split(df$size, df$type)
boxplot(l, main="file size by type", las=2)
```

```{r}
df_tcgabiolinks <- df
```

```{r}
print("time taken to pull clinical data from a single TCGA project")
print(system.time({
  clinical_luad <- GDCquery_clinic(project = "TCGA-LUAD", type = "clinical")
    
  clinical_hnsc <- GDCquery_clinic(project = "TCGA-HNSC", type = "clinical")
}))
```

This means that I can't simply merge the tables with rbind

```{r}
table(colnames(clinical_hnsc) %in% colnames(clinical_luad))
```

# cbioportalR

```{r }
library(cbioportalR)

set_cbioportal_db("public")

studies <- available_studies()

dim(studies)
```

```{r}
studies %>%  as.data.frame() |> gt() |> opt_stylize(style = 6, color = 'gray') 
```

There are 31 studies.

```{r}
table(grepl("_tcga_pan_can_atlas_2018", studies$studyId))
```

```{r }
studies_tcga <- studies$studyId[grepl("_tcga_pan_can_atlas_2018", studies$studyId)]

studies_tcga
```

## get_fusions_by_study: LUAD

```{r }
print("time taken to get fusions from a single study")
print(system.time({
  luad <- get_fusions_by_study(study_id = "luad_tcga_pan_can_atlas_2018")
}))
```

This is the type of information that we have access using the cbioportalR::get_fusions_by_study function.

```{r }
luad %>% head() %>%  as.data.frame() |> gt() |> opt_stylize(style = 6, color = 'gray') 
```

split fusions by patient

```{r }
l_patient <- split(luad$sampleId, luad$patientId)
```

Patient `r names(head(sort(sapply(l_patient, length), decreasing = T), n=1))` has the most samples with `r unname(head(sort(sapply(l_patient, length), decreasing = T), n=1))`

```{r }
head(sort(sapply(l_patient, length), decreasing = T))
```

```{r cache=TRUE}
#| label: gather_fusion_data

l_all <- list()

n <- 1
n_files <- length(studies_tcga)

print("total time needed to gather all TCGA fusion information")
print(system.time({
  for(study in studies_tcga){
  
  print_progress <- paste(date(), "working on:", study, paste(n, n_files, sep = "/"), sep = " ")
  print(print_progress)
  print(system.time({
    l_all[[study]] <- get_fusions_by_study(study_id = study)
  }))
  n <- n + 1

}
}))

# merge into a single data frame
df_all <- do.call(rbind, l_all)
```

```{r }
# out_file <- "~/long-term-scratch/bbermude/data/OMICS/Y90_2409_JRBB_Test/Data/TCGA/cbioportalR_fusions.RDS"
# saveRDS(object = l_all, file = out_file)

out_file <- "~/long-term-scratch/bbermude/data/OMICS/Y90_2409_JRBB_Test/Data/TCGA/cbioportalR_TCGA_fusions.csv"
write.csv(x = df_all, file = out_file, row.names = F)
```

```{r}
df <- df_all
```

Gathering all clinical data for all TCGA projects for which we have fusion genes data

```{r}
tcga_ids <- paste("TCGA", stringr::str_to_upper(sub("_.+", "", studies_tcga)), sep = "-")

# manual curation of some entries
tcga_ids[tcga_ids == "TCGA-COADREAD"] <- "TCGA-COAD"
tcga_ids <- c(tcga_ids, "TCGA-READ")


```

```{r }
#| label: gather_clinical_data

out_dir <- "~/long-term-scratch/bbermude/data/OMICS/Y90_2409_JRBB_Test/Data/TCGA/clinical"
dir.create(out_dir, showWarnings = F, recursive = T)
setwd(out_dir)

l_clinic <- list()

n <- 1
n_files <- length(tcga_ids)

print("total time spent getting clinical information")
print(system.time({
  for(tcga_id in tcga_ids){
      print_progress <- paste(date(), "working on:", tcga_id, paste(n, n_files, sep = "/"), sep = " ")
      print(print_progress)
      print(system.time({
        l_clinic[[tcga_id]] <- GDCquery_clinic(project = tcga_id, type = "clinical", save.csv = T)
      }))
      n <- n + 1
  }
}))
```

```{r}
hist(sapply(l_clinic, ncol))
```

```{r}
sort(sapply(l_clinic, ncol))
```

```{r}
common_cols <- Reduce(
  intersect,
  lapply(l_clinic, colnames)
)

common_cols
```

Merge all clinic information with shared columns

```{r}
df_clinic <- do.call(rbind, lapply(l_clinic, function(x) x[,common_cols]))

out_file <- "~/long-term-scratch/bbermude/data/OMICS/Y90_2409_JRBB_Test/Data/TCGA/clinical/clinical_shared_columns.csv"
write.csv(x = df_clinic, file = out_file, row.names = F)
```

```{r}
hist(df_clinic$age_at_index, main = "Age at index distribution", xlab = "Age at index")
```

# Memory usage

This is the total memory used summing all object sizes at the end of the markdown. This doesn't includes those objects that may have been deleted in different chunks of the code.

::: {.callout-note collapse="true"}
```{r}
#| label: mem-usage
# Function to convert bytes to human-readable format
human_readable <- function(bytes) {
  units <- c("B", "KB", "MB", "GB", "TB")
  if (bytes == 0) return("0 B")
  i <- floor(log(bytes, 1024))
  size <- bytes / (1024^i)
  sprintf("%.2f %s", size, units[i + 1])
}

# Get total size of all objects in the global environment
total_size <- sum(sapply(ls(envir = .GlobalEnv), function(x) object.size(get(x, envir = .GlobalEnv))))

# Print in human-readable format
cat("Total memory used:", human_readable(total_size), "\n")

```
:::

# Session Info

::: {.callout-note collapse="true"}
```{r}
#| label: Session-Info
sessioninfo::session_info()
```
:::
